Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2022-01-31 response.txt

2. At 1.46-1.70, the introduction seems to set up the F statistics as a preferred alternative to methods such as PCA and structure. However, this reasoning is unconvincing, as the various methods are all useful methods that can contribute in different ways. You could simply say that PCA is used to accomplish task X, and F statistics are used to accomplish task Y. It is motivation enough that one wants to understand how these different methods relate to each other.
This section seems to also comment on the use of F statistics to build models with large numbers of populations from calculations with 3 and 4 populations, as a contrast to (and possibly preferable to?) coalescent-style models. However, the paper does not really have much to say about these topics, and if anything, it is negative about them in the final paragraph. That makes the introduction strange in setting up F statistics as preferred to other methods but then deciding that they are problematic. I would suggest deferring any mention of the models of large numbers of populations to the Discussion. As in the previous paragraph, it’s enough to just want to understand the methods better.
RESPONSE:
I can see your point. My intention here was to simply give a brief overview of
the different methods used, and explain why one typically combines multiple
approaches by giving their main benefits and drawbacks. I can see how this is
not entirely relevant, and shortened this paragraph to remove the
benefits/drawback part.

3. In Fig 1 caption, please comment on the case in which the true relationship for four populations is ((X1,X2),(X3,X4)). Is it always assumed that species must have a relationship (((X1,X2),X3),X4) so that the symmetric case is never considered? 
RESPONSE:
Since F-statistics are distances, they correspond to unrooted trees, i.e. the
rooting is arbitrary. I changed the caption to "using an (arbitrarily rooted)
tree".

4. The paper is somewhat missing a grand summary that clarifies its main contributions. It seems that F3 and F4 computations are sometimes interpreted, problematically, as definitive tests of admixture. The paper shows that they may be redundant with PCA, or they may represent tests of particular hypotheses about placement of points in a PCA plot. I would be interested to see a table that describes two columns e.g. “ways F tests are interpreted in the literature” and “ways the geometric relationship to PCA suggests F tests should be interpreted.” We need some clarity on the meaning and appropriate use of these methods, and this paper has potential to provide a nice summary.
Some reviewers of empirical papers in human genomics reflexively expect authors to augment studies that interpret population relationships using PCA, MDS, Structure, Admixture, etc. with F tests. I gather from this paper that F tests are accomplishing the same kind of post-processing of a PCA plot as, say, the so-called average linkage distance (e.g. Kopelman et al. 2009 BMC Genetics for an example). It would be great if the paper could clarify.
I added a short conclusion where I summarize the main points I think the paper
contributes. I agree that it would be nice (and practical) to have a table
describing further applications, but an issue is that I still don't fully understand
some of the more complex of Nicks tools (such as qpgraph and qpwave), and so I do not think I can critisize
them in a constructive way.

5. In Section 5.3, the first three paragraphs seem to be setting up a significant and strong conclusion, but the last two paragraphs do not really deliver one. Move the discussion of technical consequences of complexities in human population structure for the qp methods earlier. The two paragraphs about Lewontin’s paper are not really used to make the subsequent points; authors for the special issue should not feel obligated to extensively discuss Lewontin 1972 if the relationships to their papers are tenuous.
I did rephrase this a little bit, since I agree that the criticism of qpXXX is
not warranted. I think the point would really be that F-stats and PCA could
allow us to study population structure by setting up orthogonality tests without
making treeness assumptions.

@@

131 The journal will likely insist that any section that has subsections needs to have two subsections. At present, 2.1 has only one.

138 Clarify that each SNP contributes a dimension. I do not see the high dimensionality as a drawback. The geometric framework is very intuitive, even if it involves many dimensions. Rather than framing the high dimension as a weakness, frame the dimensionality reduction as a way to make the interpretations more intuitive.
See response to reviewer 1, comment 7 who had an analogous comment

148 Sentence needs some references. These subtle details are important to anyone who wants to run a PCA, and it would be helpful to list the main places where the differences are discussed.
Added references to Cavalli-Sforza/Patterson et al. 2006. I am not sure I can
exhaustively list references, since many packages exist, and many packages have
multiple options for normalizations or dealing with missing data.
178 This connection between PCA and MDS is not understandable enough unless one already knows it – please clarify.
I decided not to re-derive this result (see also response to reviewer 1 in
previous review), and merely state the result. I rewrote this section to better
reflect that I state a known result
Page 7 Second line of eq. 10 is an approximation.
Changed the typesetting to better reflect that this is the definition of the
approximation
237 By “supposed” do you mean “assumed”? Clarify.
changed to "expected"
238 Explain why being estimates may make the values negative.
changed to "sampling noise might make some of them slightly negative"

273-282 How about a schematic figure here.
I did add a small figure (now Figure 3). Due to the difficulty of visualizing in
3D, I opted to only do a 2D plot (the logic transfers)

@@

Typo in denominator of equation 14: X3-X4 not X3-X3.
Fixed

355 Explain the significance of this holding for any orthonormal projection.
This is done in the follow-up section: This is useful in particular (...)

362-366 This text seems to be arguing for a non-identifiability of the admixture proportion from the position of the points if not collinear.
The point here is more that collinearity is an implicit assumption in admixture
proportion inference

394-397 Explain. Why are these concerns rendered unimportant?
Changed this section to explain that this is because I just aim at simplifying
the geometry, rather than interpret it in terms of biological processes

@@

404 I don’t quite follow – why is this the expectation?
Changed to: Thus, I would also expect individual-based analyses and $F_4$-statistics between closely related populations to require substantially more PCs as there local differentiation, which is typically reflected on higher PCs, becomes important.

430 What might this PCA-based F statistics framework look like?
TODO

472 These is an odd group of citations; I would use Cavalli et al. 1994 and the Jobling et al. book detailed discussions of population history. Since classic apportionment does not give us a way to think about ancient DNA, you’d want to refer to Witt et al. in the special issue if you are wanting to connect Lewontin’s paper to ancient DNA such as in Reich 2018.
482 Cite Novembre 2022 here.
I added a tentative entry, that I will supplement as soon as I have full
bibliographic info.

@@

Doc:
-Throughout the document, the distinction between the expectations of the statistics (F2, F3, F4 in Patterson notation) and the statistics themselves (f2, f3, f4 in Patterson notation) is mostly avoided. Overall, I think it is the right decision given the manuscript's focus, but in some places it could lead to confusion. As one example, in line 71-73, the claim about F3 being restricted to be non-negative under a tree structure applies to F3 but not necessarily f3. Please address this distinction somewhere (presumably in the intro or theory sections, up to your discretion) and explain it in relation to the manuscript's setting in allele-frequency space.
Added a paragraph after the definition of F-statistics:
Commonly, a distinction is made between statistics estimated from data (denoted with lowercase-$f$), and theoretical quantities (defined in eq. 1). I do not make this distinction, but will explicitly mention when I analyze statistics calculated from data.

-lines 259-260 - this point is phrased somewhat vaguely. Please consider a more precise phrasing.
changed to: "In converse, populations that are distant on the first few PCs are
guaranteed to also have a large $F_2$-distance, since the distance contributed
from the omitted PCs cannot be negative."

-lines 273-277 - The globe analogy is useful, but it needs to be explicitly connected to the problem. (For example, state that in the example, the north and south poles represent population allele frequencies, the globe represents the space in which F3<=0, etc.) Another schematic of this scenario, though perhaps challenging to draw, would be useful, for example with two points in 3D space defining a ball, two orthogonal slices representing projections, and the circles (in 2d) defined by the positions of the points in the projections.

-In the same section about F3 on a 2D PCA plot, please explain how to compute the true (light gray in figure 3) and apparent (dark gray in figure 3) circles. The intuition for the dark gray circle comes across via figure 2, but it will not be obvious to many readers how to think of the light circle. A rigorous calculation may not be necessary. One intuitive explanation (if the following is a correct reading!) might be that one computes the distance between the populations in the full space, then draws a circle with that diameter centered on the midpoint between the positions of the points in the projection. It would be helpful to contrast this with an explanation of the dark grey circle as centered in the same spot, but with a diameter defined by the distance between the points in the current 2D projection, which is guaranteed to be no greater than the full distance.

-In your eq A4, plus lines 278-282, you note that there is a strict interpretation that if the point lies outside the circle in any one projection, then it is guaranteed to be outside the n-ball. You also state that it is possible for it to lie inside the circle on any one biplot yet still have positive F3. I think it is also true that it can lie inside the (light gray) circle on all biplots and still have positive F3. (e.g. in 3D, the volume contained in the intersection of the cylinders defined by the circles from the 2D projections will be a bicylinder that completely contains the ball and has a bit of extra volume too.) Please address this.
(Response to last three comments): I agree that this is an important issue, and
now explicitly addressed in Figure 3. Unfortunately, I could not make this work
in 3D, but I think the 2D->1D Figure also works well, as the intuition
translates.

### 2022-01-31 fstats_pca_lewontin.tex
merge

\subsection{$F_2$ in PC-space}
The $F_2$-statistic is an estimate of the squared allele-frequency distance between two
populations. On a tree (Figure \ref{fig:geom}A) this corresponds to the branch between two populations (Figure \ref{fig:geom}E). In allele-frequency space, it corresponds to the squared Euclidean distance, and thus reflects the intuition that closely related populations  will fall close to each other on a PCA-plot, and have low pairwise $F_2$-statistics. However, since $F_2$ can be written as  a sum of squared (non-negative) terms for each PC (eq. \ref{eq:fpc}), the distance on a PCA-plot will always be an underestimate of the full $F_2$-distance. Thus, PCA might project two populations with high $F_2$-distance very close to each other, which would indicate that these particular PCs are not suitable to  understand and visualize the relationship between these particular populations, and likely more PCs need to be investigated to understand how these populations are related to each other. In converse, populations that are distant on a PCA-plot are guaranteed to also have a large $F_2$-distance.

\begin{figure}[!ht]
	\includegraphics[width=.4\textwidth]{figures/fig_proj.pdf}
	\caption{\textbf{Admixture-$F_3$-statistic and Dimensionality reduction.} A: $F_3(X_x; X_1, X_2)$ in two dimensions (full data). Points like $X_x$ inside the circle (red) have $F_3 <0$, but points outside ($X_3, X_4$, blue) have positive $F_3$-statistics. B: Projection on one dimension. The full rejection region (light gray bar) is larger than that predicted from the distance between $X_1$ and $X_2$ (dark grey bar). Points with negative $F_3$ are guaranteed to lie in the light gray (but not dark gray). $X_3$ also projects onto the gray bar, even though its $F_3$ is positive. Points such as $X_4$ that project outside the rejection region have $F_3>0$.}
	\label{fig:proj}
\end{figure}

@@

\subsection{The Geometry of Admixture}
Previous interpretation of PCA in the context of population genetic models have focused on explicit models and aimed at directly interpreting the PCs in terms of population genetic parameters \citep{cavalli-sforza1975, novembre2008a, francois2010, francois2021}. My interpretation here is different in that the utility of PCA is to simplify the geometry of the data. One consequence is that the results here are not impacted by sample ascertainment, or sample sizes which are common concerns in the interpretation of PCA \citep{novembre2008a, mcvean2009, francois2010}. Adding more PCs will provide a gradually better approximation of $F$-statistics, and a  skewed sampling distribution would likely require more PCs to be included in an analysis.

@@

However, understanding, conceptualizing and modeling this variation is far from trivial, particularly in a historical context in which mistaken ideas about human variation have been used to justify racist, eugenic and genocidal policies. Lewontin’s landmark 1972 paper on the apportionment of human genetic diversity was one of the first to quantify how little of between-population genetic variation could be attributed to ``racial'' continental-scale groupings. Over the last five decades, this view has been corroborated, refined and extended many times \citep{cann1987, cavalli-sforza1994, barbujani1997, rosenberg2002a}.

@@

This continuum between admixture and population structure becomes more problematic when interpreting tools that  use multiple $F$-statistics to build compound models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as I do not need to assume that gene flow is rare.

\section{Conclusion}
\begin{itemize}
	\item $F$-stats and PCA are partially redundant
	\item Population structure is sparse - trees	
	\item Formulating hypotheses
	\item Orthogonality as a central concept
	\item Orthogonality of scales (within-population variation) is independent of between-population variation
	\item Orthogonality of isolation (two locations have little exchange)
\end{itemize}
\section{Supplementary File Description}
Both supplementary files give the individual-identifier, sex and population label for each sample.

>>>>

\subsection{$F_2$ in PC-space}
The $F_2$-statistic is an estimate of the squared allele-frequency distance between two
populations. On a tree (Figure \ref{fig:geom}A) this corresponds to the branch between two populations (Figure \ref{fig:geom}E). In allele-frequency space, it corresponds to the squared Euclidean distance, and thus reflects the intuition that closely related populations  will fall close to each other on a PCA-plot, and have low pairwise $F_2$-statistics. However, since $F_2$ can be written as  a sum of squared (non-negative) terms for each PC (eq. \ref{eq:fpc}), the distance on a PCA-plot will always be an underestimate of the full $F_2$-distance. Thus, PCA might project two populations with high $F_2$-distance very close to each other, which would indicate that these particular PCs are not suitable to  understand and visualize the relationship between these particular populations, and likely more PCs need to be investigated to understand how these populations are related to each other. In converse, populations that are distant on the first few PCs are guaranteed to also have a large $F_2$-distance, since the distance contributed from the omitted PCs cannot be negative.

\begin{figure}[!ht]
	\includegraphics[width=.4\textwidth]{figures/fig_proj.pdf}
	\caption{\textbf{Admixture-$F_3$-statistic and Dimensionality reduction.} A: The gray circle is defined by $X_1$ and $X_2$ and reflects the area for which $F_3$ is negative, i.e. $F_3(X_x; X_1, X_2) < 0 < F_3(X_3; X_1, X_2) \leq F_3(X_4; X_1, X_2)$. B: Projection on one dimension. The full rejection region (light gray bar) is given by the boundaries of the circle in A, and is larger than that predicted from the distance between $X_1$ and $X_2$ (dark grey bar) on PC1 alone. Points like $X_x$ with negative $F_3$ are guaranteed to lie in the light gray (but not dark gray) region. $X_3$ also projects onto the gray bar, even though it is outside the circle and $F_3$ is positive, and it would also lie inside the positive area if projected onto PC2. However, points such as $X_4$ that project outside the rejection region have $F_3>0$, since the associated projection line does not intercept the circle.}
	\label{fig:proj}
\end{figure}

@@

\subsection{The Geometry of Admixture}
Previous interpretation of PCA in the context of population genetic models have focused on explicit models and aimed at directly interpreting the PCs in terms of population genetic parameters \citep{cavalli-sforza1975, novembre2008a, francois2010, francois2021}. My interpretation here is different in that the utility of PCA is to simplify the geometry of the data, rather than attributing meaning to the produced PCs. One consequence is that the results here are less directly impacted by sample ascertainment, sample sizes or number-of-PCs, which are common concerns in the interpretation of PCA \citep{novembre2008a, mcvean2009, francois2010, francois2021}; adding more PCs will provide a successively better approximation of the $F$-statistics.

@@

However, understanding, conceptualizing and modeling this variation is far from trivial, particularly in a historical context in which mistaken ideas about human variation have been used to justify racist, eugenic and genocidal policies. Lewontin’s landmark 1972 paper on the apportionment of human genetic diversity was one of the first to quantify how little of between-population genetic variation could be attributed to ``racial'' continental-scale groupings \citep{novembre2022}. Over the last five decades, this view has been corroborated, refined and extended many times \citep{cann1987, cavalli-sforza1994, barbujani1997, rosenberg2002a}.

@@

\begin{comment}
This continuum between admixture and population structure becomes more problematic when interpreting tools that  use multiple $F$-statistics to build compound models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as I do not need to assume that gene flow is rare.
\end{comment}

An alternative could be to focus on orthogonality of drift as a \emph{the} central concept for analyzing genetic variation data.
This is often informally done in the context of PCA, where it is often observed that a particular PC is associated with variation within a specific region, or separates out distinct populations \citep{cavalli-sforza1994}. These associations are not exact, and impacted e.g. by the size and composition of the analyzed data set. On the other hand, we can use $F$-statistics to formally test orthogonality. Distinct clades in an admixture graph would be one possibility how orthogonality could arise, but assuming a graph is not necessary. Motivated by PCA, we could set up $F_4$ as tests of orthogonality due to either space (distinct populations evolve independently) or scale (i.e. between-population diversity is independent of within-population diversity). This slight generalization of $F$-statistics could allow us to reframe many of the questions that are currently asked as tests of orthogonality without assuming discrete lineages.

\section{Conclusion}
$F$-statistics and PCA have both proven to be tremendously useful to study, visualize and test aspects of population structure. Here, I show that these approaches are closely connected, and highlight a few implications. First, PCA and $F$-statistics should never be treated as independent analyses. If they agree, this can be used as a sanity check that no major assumptions about e.g. population groupings are violated, but the underlying biological relationships investigated are the same in both approaches.

If sufficient PCs are considered, both $F_3$ and $F_4$-statistics do have simple interpretations in PCA-space. If used as a test for admixture, $F_3$ corresponds to a $n$-sphere in PCA-space, which manifests as a circle on a PCA-plot. The informal notion that an admixed population should lie between its sources on a PCA-plot is necessary, but not sufficient; unadmixed populations may also project there. In contrast, it is sufficient (but not necessary) to determine that a population has a positive $F_3$ statistic.

Interpretations of the outgroup-$F_3$-statistics and $F_4$-statistics in a PCA-framework rely heavily on the idea of a projection: In a tree or admixture-graph framework, we can loosely interpret $F_4(A, B; C, D)$ as a projection of one the path $A-B$ onto $C-D$; only edges shared between the paths will contribute non-zero amounts to this statistics. This projection analogy holds when we replace the populations by points in PCA-space, which enables us to study population models beyond discrete graphs. In this context, $F_4$ is simply an orthogonality test, an $F_4$ of zero means that the pairs of populations are orthogonal (i.e. independent) of another.

One difficulty in interpreting the results here is that they only hold when all PCs are considered. Yet, in my applications I found that two (for the World data set) and three PCs (for Western Eurasia) already provided a very good approximation for the majority of $F$-statistics, so that the results here should be applicable to at least continental-scale data sets.

\section{Supplementary File Description}
Both supplementary files give the individual-identifier, sex and population label for each sample.
